F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-23 icfp-2015/paper.tex
Adding more stuff to semantics

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string

@@

@@

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}

@@

\authorinfo{Tomas Petricek}
           {University of Cambridge}
           {tomas@tomasp.net}
%\authorinfo{Name2\and Name3}
%           {Affiliation2/3}
%           {Email2/3}

\maketitle

@@

\begin{figure*}
\begin{equation*}
\begin{array}{l}
 \ident{asInt}(i) \,\reduce i\\
 \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
 \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
 \ident{asInt}(t) \,\reduce i\quad(t~\text{\small represents}~i) \\
 \ident{asInt}(\kvd{true}) \reduce 1\\
 \ident{asInt}(\kvd{false}) \reduce 0\\
\end{array}
\;
\begin{array}{l}
 \ident{asDec}(i) \,\reduce d\quad(d = i)\\
 \ident{asDec}(d) \reduce d\\
 \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
 \ident{asDec}(t) \,\reduce d\quad(t~\text{\small represents}~d) \\
 \ident{asDec}(\kvd{true}) \reduce 1.0\\
 \ident{asDec}(\kvd{false}) \reduce 0.0\\
\end{array}
\;
\begin{array}{l}
 \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
 \ident{asFloat}(d) \reduce f\quad(f = d)\\
 \ident{asFloat}(f) \reduce f\\
 \ident{asFloat}(t) \,\reduce f\quad(t~\text{\small represents}~f) \\
 \ident{asFloat}(\kvd{true}) \reduce 1.0\\
 \ident{asFloat}(\kvd{false}) \reduce 0.0\\
\end{array}
\;
\begin{array}{l}
 \ident{asStr}(i) \,\reduce t\quad(t~\text{\small represents}~i)\\
 \ident{asStr}(d) \reduce t\quad(t~\text{\small represents}~d)\\
 \ident{asStr}(f) \reduce t\quad(t~\text{\small represents}~f)\\
 \ident{asStr}(t) \,\reduce t\\
 \ident{asStr}(\kvd{true}) \reduce \str{true}\\
 \ident{asStr}(\kvd{false}) \reduce \str{false}\\
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-1.5em}
\begin{array}{l}
 \ident{asBool}(1) \reduce \kvd{true}\\
 \ident{asBool}(1.0) \reduce \kvd{true}\\
 \ident{asBool}(\str{true}) \reduce \kvd{true}\\
 \ident{asBool}(\str{1}) \reduce \kvd{true}
 \\[0.6em]
 \ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
 \ident{isNull}(\_) \reduce \kvd{false} 
\end{array}
\hspace{2.6em}
\begin{array}{l}
 \ident{asBool}(0) \reduce \kvd{false}\\
 \ident{asBool}(0.0) \reduce \kvd{false}\\
 \ident{asBool}(\str{false}) \reduce \kvd{false}\\
 \ident{asBool}(\str{0}) \reduce \kvd{false}\\
 \ident{asBool}(b) \reduce b
 \\[0.6em]
 \ident{getChildren}(\kvd{null}) \reduce [~]
\end{array}
\hspace{3.0em}
\begin{array}{l}
 \ident{getField}_{\nu,\nu_i}(\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
 \ident{getField}_{\bot, \nu_i}(\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
 \ident{getField}_{\nu,\nu'}(\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
 \ident{getField}_{\bot, \nu'}(\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
 \\[0.6em]
 \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-1.5em}
\begin{array}{l}
 \ident{isTag}_\ident{string}(t) \reduce \kvd{true} \\
 \ident{isTag}_\ident{bool}(v) \reduce \kvd{true} \quad(\textnormal{when}~v\in{0,1,0.0, 1.0, \kvd{true},\kvd{false}} )\\
 \ident{isTag}_\ident{number}(v) \reduce \kvd{true} \quad(\textnormal{when}~v=i, v=d, v=f) \\
 \ident{isTag}_\ident{collection}([s_1; \ldots; s_n]) \reduce \kvd{true} \\
\end{array}
\hspace{2.5em}
\begin{array}{l}
 \ident{isTag}_\ident{collection}(\kvd{null}) \reduce \kvd{true} \\
 \ident{isTag}_\ident{rec-anon}(\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
 \ident{isTag}_{\ident{rec-named}~\nu} (\nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
 \ident{isTag}_{\_}(\_) \reduce \kvd{false} \\
\end{array}
\end{equation*}

@@

As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). To capture these, the operational semantics accesses 
values using the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s) \lsep \ident{asStr}(s) \\
    &\narrow{|}& \ident{getChildren}(s) \lsep \ident{getField}_{\nu_{\ident{opt}}, \nu}(s) \lsep \ident{isNull}(s) \lsep \ident{isTag}_t(s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we intend to prove about F\# Data provides \emph{sufficient}, but
not \emph{necessary} conditions. This means that a type provided based on a sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

@@

@@

The semantics fragment introduced in the previous section discusses values and operations that are used by
F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
only need class types (without inheritance) with properties. A class has a single implicit constructor and 
the declaration closes over constructor parameters. In addition, we include special constructs for working
with option types and lists.

\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
 e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& [e_1;\ldots;e_n]\lsep\kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
\end{array}
\end{equation*}
%
In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
previously defined operations $op$, class construction, member access, conditionals and expressions for working 
with option values and lists.

@@

\vspace{-0.5em}
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& [\cdot] \lsep f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 
 \\[0.6em]
 f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag}_t \;\}
\end{array} 
\end{equation*}

@@

The language presented here is very simple. In particular, it does not provide any form of recursion
and so all expressions reduce to a value in a finite number of steps or get stuck due to an error
condition. An error condition can be a wrong argument passed to conditional, pattern matching or
one of the conversion functions from Figure~\ref{fig:op-conversions}.

@@

\begin{figure}
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
 \inference
  { e \reduce e' }
  { E[e] \reduce E[e']}\\
 \\
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.4em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})=\ldots \kvd{member}~N_i : \tau_i = e_i \ldots }
 { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
  [x_1;\ldots;x_m ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\quad(m\neq n) \\
 \\
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
  [x_1;\ldots;x_n ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
\end{array}
\end{equation*}

\caption{Remaining reduction rules for Featherweight F\#}
\label{fig:ff-reduction}
\end{figure}

\begin{figure*}
\begin{multicols}{2}
% Primitive structural types become corresponding F# types and conversion is inserted
\begin{equation*}
\begin{array}{l}
 \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.6em]
\quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
\hspace{2.9em} (\ident{int}, \ident{int}, \ident{asInt}), (\ident{decimal},\ident{decimal},\ident{asDec}),\\[0.2em]
\hspace{2.9em} (\ident{float},\ident{float},\ident{asFloat}), (\ident{string},\ident{string},\ident{asStr}) ~\}\\[0.6em]
\end{array}
\end{equation*}
%
% Records become classes
\begin{equation*}
\begin{array}{l}
 \sem{\nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
 \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
 \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
 \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n  \\[0.1em]
 \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
 \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}_{\nu_{\ident{opt}}, \nu_i}(e)\\[0.6em]
\end{array}
\end{equation*}
%
% Collections
\begin{equation*}
\begin{array}{l}
 \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{getChildren}(e'), L \;\;\textnormal{where}~ \tau, e', L = \sem{\hat{\sigma}}_e
\end{array}
\end{equation*}

% Option values
\begin{equation*}
\begin{array}{l}
 \sem{\hat{\sigma}\;\,\kvd{option}}_e = \\[0.2em]
 \hspace{1.25em} \ident{option}\langl\tau\rangl, \kvd{if}~\ident{isNull}~e'~\kvd{then}~\ident{None}~\kvd{else}~\ident{Some}(e'), L\\[0.2em] 
 \hspace{1.25em} \textnormal{where}~\tau, e', L = \sem{\hat{\sigma}}_e
\end{array}
\end{equation*}
%
% Sum type
\begin{equation*}
\begin{array}{l}
 \sem{\sigma_1 + \ldots + \sigma_n}_e = \\[0.1em]
 \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
 \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
 \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
 \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=e_i'\\[0.1em]
 \qquad \;\,\,e_i' = \kvd{if}~\ident{isNull}(e_i)~\kvd{then}~\ident{None}~\kvd{else} \\
 \hspace{4.8em}     \kvd{if}~\ident{isTag}_{t_i}{(e_i)}~\kvd{then}~
     \ident{Some}(e_1)~\kvd{else}~\ident{None} \\[0.1em]
 \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
\end{array}
\end{equation*}
%
% Anything else
\begin{equation*}
\begin{array}{l}
 \sem{\top}_v = \sem{\kvd{null}}_v = \ident{StructVal}, v, \emptyset
\end{array}
\end{equation*}
\end{multicols}

>>>>

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit

@@

\newcommand{\llangl}{\langl\hspace{-0.35em}\langl}
\newcommand{\rrangl}{\rangl\hspace{-0.35em}\rangl}

@@

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semalt}[1]{\llangl #1 \rrangl}

@@

\authorinfo{Tomas Petricek}
           {University of Cambridge}
           {tomas@tomasp.net}
\authorinfo{Gustavo Guerra}
           {Microsoft Corporation, London}
           {gustavo@codebeside.org}
\authorinfo{Don Syme}
           {Microsoft Research, Cambridge}
           {dsyme@microsoft.com}
\maketitle

@@

\begin{figure*}
\begin{equation*}
\begin{array}{l}
 \ident{asInt}(i) \,\reduce i\\
 \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
 \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
 \ident{asInt}(t) \,\reduce i\quad(t~\text{\small represents}~i) \\
 \ident{asInt}(\kvd{true}) \reduce 1\\
 \ident{asInt}(\kvd{false}) \reduce 0\\
\end{array}
\;
\begin{array}{l}
 \ident{asDec}(i) \,\reduce d\quad(d = i)\\
 \ident{asDec}(d) \reduce d\\
 \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
 \ident{asDec}(t) \,\reduce d\quad(t~\text{\small represents}~d) \\
 \ident{asDec}(\kvd{true}) \reduce 1.0\\
 \ident{asDec}(\kvd{false}) \reduce 0.0\\
\end{array}
\;
\begin{array}{l}
 \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
 \ident{asFloat}(d) \reduce f\quad(f = d)\\
 \ident{asFloat}(f) \reduce f\\
 \ident{asFloat}(t) \,\reduce f\quad(t~\text{\small represents}~f) \\
 \ident{asFloat}(\kvd{true}) \reduce 1.0\\
 \ident{asFloat}(\kvd{false}) \reduce 0.0\\
\end{array}
\;
\begin{array}{l}
 \ident{asStr}(i) \,\reduce t\quad(t~\text{\small represents}~i)\\
 \ident{asStr}(d) \reduce t\quad(t~\text{\small represents}~d)\\
 \ident{asStr}(f) \reduce t\quad(t~\text{\small represents}~f)\\
 \ident{asStr}(t) \,\reduce t\\
 \ident{asStr}(\kvd{true}) \reduce \str{true}\\
 \ident{asStr}(\kvd{false}) \reduce \str{false}\\
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-1.5em}
\begin{array}{l}
 \ident{asBool}(1) \reduce \kvd{true}\\
 \ident{asBool}(1.0) \reduce \kvd{true}\\
 \ident{asBool}(\str{true}) \reduce \kvd{true}\\
 \ident{asBool}(\str{1}) \reduce \kvd{true}
 \\[0.6em]
 \ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
 \ident{isNull}(\_) \reduce \kvd{false} 
\end{array}
\hspace{2.6em}
\begin{array}{l}
 \ident{asBool}(0) \reduce \kvd{false}\\
 \ident{asBool}(0.0) \reduce \kvd{false}\\
 \ident{asBool}(\str{false}) \reduce \kvd{false}\\
 \ident{asBool}(\str{0}) \reduce \kvd{false}\\
 \ident{asBool}(b) \reduce b
 \\[0.6em]
 \ident{getChildren}(\kvd{null}) \reduce [~]
\end{array}
\hspace{3.0em}
\begin{array}{l}
 \ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
 \ident{getField}(\bot, \nu_i, \{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
 \ident{getField}(\nu,\nu', \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
 \ident{getField}(\bot, \nu', \{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
 \\[0.6em]
 \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-1.5em}
\begin{array}{l}
 \ident{isTag}(\ident{string}, t) \reduce \kvd{true} \\
 \ident{isTag}(\ident{bool}, v) \reduce \kvd{true} \quad(\textnormal{when}~v\in{0,1,0.0, 1.0, \kvd{true},\kvd{false}} )\\
 \ident{isTag}(\ident{number}, v) \reduce \kvd{true} \quad(\textnormal{when}~v=i, v=d, v=f) \\
 \ident{isTag}(\ident{collection}, [s_1; \ldots; s_n]) \reduce \kvd{true} \\
\end{array}
\hspace{1.8em}
\begin{array}{l}
 \ident{isTag}(\ident{collection}, \kvd{null}) \reduce \kvd{true} \\
 \ident{isTag}(\ident{rec-anon}, \{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
 \ident{isTag}(\ident{rec-named}~\nu, \nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
 \ident{isTag}(\_, \_) \reduce \kvd{false} \\
\end{array}
\end{equation*}

@@

As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(t, s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

@@

\begin{figure}
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.4em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})=\ldots \kvd{member}~N_i : \tau_i = e_i \ldots }
 { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
 \\
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
  [x_1;\ldots;x_m ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\quad(m\neq n) \\
 \\
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
  [x_1;\ldots;x_n ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\
 \textnormal{\footnotesize{(map)}}&
 \hspace{-0.4em}
 \ident{List.map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
\end{array}
\end{equation*}

\caption{Remaining reduction rules for Featherweight F\#}
\label{fig:ff-reduction}
\vspace{-0.5em}
\end{figure}

@@

The semantics fragment introduced in the previous section discusses values and operations that are used by
F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
only need class types (without inheritance) with properties. A class has a single implicit constructor and 
the declaration closes over constructor parameters. In addition, we include special constructs for working
with option types and lists.
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
 e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
\end{array}
\end{equation*}

\noindent
In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
previously defined operations $op$, class construction, member access, conditionals and expressions for working 
with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.

@@

\noindent
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep [\cdot] 
 \\[0.6em]
 f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \;\}
\end{array} 
\end{equation*}

@@

The language is sufficient for our purpose, but simple. It does not provide recursion
and so all expressions reduce to a value in a finite number of steps or get stuck due to an error
condition. An error condition can be a wrong argument passed to conditional, pattern matching or
one of the conversion functions from Figure~\ref{fig:op-conversions}.

@@

\begin{figure*}
\begin{multicols}{2}
% Primitive structural types become corresponding F# types and conversion is inserted
\noindent
\begin{equation*}
\begin{array}{l}
 \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.4em]
\quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
\hspace{2.9em} (\ident{int}, \ident{int}, \ident{asInt}), (\ident{decimal},\ident{decimal},\ident{asDec}),\\[0.2em]
\hspace{2.9em} (\ident{float},\ident{float},\ident{asFloat}), (\ident{string},\ident{string},\ident{asStr}) ~\}\\[0.6em]
\end{array}
\end{equation*}
%
% Records become classes
\begin{equation*}
\begin{array}{l}
 \sem{\nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
 \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
 \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
 \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n  \\[0.1em]
 \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
 \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu_{\ident{opt}}, \nu_i, e)\\[0.6em]
\end{array}
\end{equation*}
%
% Collections
\begin{equation*}
\begin{array}{l}
 \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)), L\\[0.4em]
 \quad \textnormal{where}~~\tau, e', L = \sem{\hat{\sigma}}_x
\end{array}
\end{equation*}

% Option values
\noindent
\begin{equation*}
\begin{array}{l}
 \sem{\hat{\sigma}\;\,\kvd{option}}_e = \\[0.2em]
 \hspace{1.25em} \ident{option}\langl\tau\rangl, \kvd{if}~\ident{isNull}~e~\kvd{then}~\ident{None}~\kvd{else}~\ident{Some}(e'), L\\[0.2em] 
 \hspace{1.25em} \textnormal{where}~\tau, e', L = \sem{\hat{\sigma}}_e
\end{array}
\end{equation*}
%
% Sum type
\begin{equation*}
\begin{array}{l}
 \sem{\sigma_1 + \ldots + \sigma_n}_e = \\[0.1em]
 \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
 \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
 \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
 \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\[0.1em]
 \hspace{5.8em}  \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\
 \hspace{5.8em}  \kvd{if}~\ident{isTag}(t_i, v)~\kvd{then}~
     \ident{Some}(e_i)~\kvd{else}~\ident{None} \\[0.1em]
 \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
\end{array}
\end{equation*}
%
% Anything else
\begin{equation*}
\begin{array}{l}
 \sem{\top}_v = \sem{\kvd{null}}_v = \ident{StructVal}, v, \emptyset
\end{array}
\end{equation*}
\end{multicols}

\vspace{-0.5em}
\noindent
\begin{equation*}
\qquad
\qquad
\begin{array}{rcl}
 \nameoftag(\ident{string}) &\narrow{=}& \ident{String} \\
 \nameoftag(\ident{bool}) &\narrow{=}& \ident{Boolean} \\
\end{array}
\qquad
\begin{array}{rcl}
 \nameoftag(\ident{number}) &\narrow{=}& \ident{Number} \\
 \nameoftag(\ident{collection}) &\narrow{=}& \ident{List} \\
\end{array}
\qquad
\begin{array}{rcl}
 \nameoftag(\ident{rec-anon}) &\narrow{=}& \ident{Record} \\
 \nameoftag(\ident{rec-named}\;\nu) &\narrow{=}& \nu \\
\end{array}
\end{equation*}

\caption{Type provider -- generation of featherweight F\# types from inferred structural types}
\label{fig:tp-generation}
\vspace{-0.5em}
\end{figure*}
